---
title: "Austin TX. Animal Shelter Data"
author: "Ricardo Mirville"
date: "2022-09-05"
output:
  html_document: default
  pdf_document: default
---

# Data Exploration with Austin TX Animal Shelter Data #

Something that I am very passionate about is animals.

Shelters are known to home many different animals that are either found wandering the streets or brought in because their owner could no longer care for them.

The *most common* animals placed in shelters are known to be cats and dogs, so in this data exploration I will be focusing and exploring the two species to see what type of insights we get.

Though there are so much we can gather from this data set alone, The Key things I think are important to note are:

* Comparison between canine and feline entering into the shelter from years 2013 - 2022
* View the top 10 breeds that are admitted into the shelter for each specific reasoning
* See if we can see trends in Ages for both Canine and Feline
* Comparison between the type of intake and their condition 

By the end of this exploration, we will get a better picture as to which top ten cat or dog breeds are known to come into the shelter and what their reasons may be.

### Objectives of Data Exploration ###

1. **Create** and establish my environment
2. **Import** the data into R Studio
3. **Clean** and organize the data
4. **Prepare** data for exploration
5. **Summarize** findings and correlation

## Establishing R Environment ##

Let's load our environment with `tidyverse` `lubridate` `dplyr` `ggplot2` 

```{r environment}
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
```



Now that we have everything we are going to need for the environment lets load the data.




## Importing Austin TX Animal Shelter Dataset ##

```{r import data}
a_df <- read_csv("/Users/ricardomirville/Desktop/Austin_Animal_Center_Intakes.csv")
```



Lets see what this data looks like and what it contains




```{r}
head(a_df)
summary(a_df)
```

## Cleaning and Organizing the Data ##


This data has a lot of great information to work with, but there are factors in the data set that I would like to clean up as well as fix class issues.


### Cleaning Objectives: ###

1. Rename columns
2. Create better names for values
3. Fix class issues


```{r renaming columns}
a_df <- a_df %>% 
  rename(Species = `Animal Type`, Date = DateTime, Sex=`Sex upon Intake`) 


# In the data set we notice that `DateTime` and `MonthYear` have duplicate values. We will remove one of these rows in the Prepare Data for Exploration phase.
```


Now that we have cleaned up some of the column names, lets address some of the values within the data set that can be changed for better readability

First in `Species` lets change cat and dog to Feline and Canine

```{r rename cat/dog}
a_df <- a_df %>% 
  mutate(Species = ifelse(Species == "Cat", "Feline", Species)) %>% 
  mutate(Species= ifelse(Species=="Dog","Canine",Species))
```

```{r}
 # Always good to check to make sure everything is correct

unique(a_df$Species)
```

If we take a look at the `Age upon Intake` column we notice there several animals that are under one years old and can be categorized as month / day/ week old. To clean this up, any animal under one year old will be classified as `<1 year`

```{r cleaning age colulmn}
#Lets see all the unique ages of all the animals
unique(a_df$Age upon Intake)
```

```{r}
# time to change all the ages that are less than one year
a_df <- a_df %>% 
  mutate(`Age upon Intake`= ifelse(`Age upon Intake` =="11 months","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake` ifelse(`Age upon Intake`=="4 weeks","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`= "5 months","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`= ifelse(`Age upon Intake`=="1 month","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="2 months","<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="4 months","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="6 months","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="4 days", "<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="1 day","<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="2 weeks","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="3 weeks", "<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="3 months", "<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="1 week", "<1 year",`Age upon Intake`) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`="9 months","<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="10 months","<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="7 months","<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="8 months","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="1 weeks", "<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="5 days", "<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="2 days","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`= ifelse(`Age upon Intake`=="0 years", "<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="3 days", "<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="5 weeks","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="6 days","<1 year",`Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="-1 years", "<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="-2 years", "<1 year", `Age upon Intake`)) %>% 
  mutate(`Age upon Intake`=ifelse(`Age upon Intake`=="-3 years", "<1 year", `Age upon Intake`))
```

```{r}
# Confirming everything has been cleaned properly

unique(a_df$`Age upon Intake`)
```


Lastly I want to change the class of `Date` from CHR to DATE as well as exclude the time.


```{r}
a_df <- a_df %>% 
  mutate(Date = substring(Date,1,10)) %>% 
  mutate(Date = mdy(Date))
```

```{r}
# Confirming we changed the class from char to date.
class(a_df$Date)
```

## Data Preperation ##

Now that we have cleaned the data, it is time to Prepare the Data for Exploration and Analysis

### Steps for Preperation ###

1. Select the columns needed
2. Filter for ONLY Feline and Canine species.

```{r Data Prep}
a_df <- a_df %>% 
  select(Date,`Intake Type`,`Intake Condition`,Species,`Age upon Intake`, Breed,Sex)
```

```{r}
a_df <- a_df %>% 
  filter(!(Species == "Other" | Species == "Livestock" | Species == "Bird"))
```

```{r}
#Double checking the code is correct.
unique(a_df$Species)
```

Perfect now our data is ready to undergo some exploratory analysis.

## Exploratory Analysis ##


First it would be important to visualize the canine and feline intake over the years of 2013-2021 as provided in the data set.

```{r}
#first we'll order the date chronologically
a2_df<- a_df[order(a_df$Date),]

#next would be to create our new table.

a2_df <- a2_df %>% 
  select(Date,Species) %>% 
  group_by(Date,Species) %>% 
  summarise(Total_Count = n())

a2_df

#Now to plot it out

```


```{r Total intake Graph}
ggplot(a2_df, aes(x=Date,y =Total_Count,col = Species,shape = Species))  +scale_color_manual(values = c("#3458da","#da3477")) +geom_point() +labs(title = "Total Intake for Canine and Feline 2013 - 2021", caption = "Dates shown from October 03, 2013 through March 03, 2021")
```


Animals have many reasons for coming into the shelter, in this case for this data set, the `Intake Type` are:

- `Stray`
- `Owner Surrender`
- `Public Assist`
- `Abandonded`
- `Euthanasia Request`

For each of these `Intake Types` we will view the **Top 10 Breeds** that are brought in to the shelter for both Canine and Feline Species.

But, it is also important to note the Top 10 Canine and Feline breeds that have entered the shelter in total from the years 2013 - 2021. Before we break down the Intake types,  we'll do that first.

To visualize this we will be using `ggplot2` and `geom_bar`

```{r}
library(ggplot2)
```



## Top 10 Canine Breeds from 2013 - 2021 ##

```{r}
C_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Canine") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount)); 

C_bar <- C_bar[1:10,];

ggplot(C_bar, aes(x= Breed, y= BreedCount)) +geom_bar(stat = "identity", fill = "dark blue", color = "black", width = .60)+ labs(title = "Top 10 Canine Breeds Admitted In The Shelter From 2013 - 2021") + theme(axis.text.x = element_text(angle = 20,size = 8))
```

## Top 10 Feline Breeds from 2013 - 2021 ##

```{r}
F_bar <- a_df%>% 
  group_by(Breed) %>% 
  filter(Species == "Feline") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

F_bar <- F_bar[1:10,];

ggplot(F_bar, aes(x= Breed,y=BreedCount)) +geom_bar(stat = "identity", fill= "pink", color = "black", width = .60) + labs(title = "Top 10 Feline Breeds Admitted In The Shelter From 2013 - 2021") + theme(axis.text.x = element_text(angle = 20,size = 8))
```


## Intake type `Stray` ##


```{r Canine Stray}
SC_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Canine") %>% 
  filter(`Intake Type` == "Stray") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

SC_bar <- SC_bar[1:10,];

ggplot(SC_bar, aes(x= Breed, y = BreedCount)) + geom_bar(stat = "identity", fill = "dark blue", color= "black", width = .60) + labs(title = "Top 10 Canine Breeds With Intake Type as Stray", caption = "From Years 2013 - 2021") + theme(axis.text.x = element_text(angle = 20, size = 8))
```


```{r Feline Stray}
SF_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Feline") %>% 
  filter(`Intake Type` == "Stray") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

SF_bar <- SF_bar[1:10,];

ggplot(SF_bar, aes(x=Breed, y= BreedCount)) + geom_bar(stat = "identity", fill = "pink", color = "black", width = .60) + labs(title = "Top 10 Feline Breeds With Intake Type as Stray", caption = "From Years 2013 - 2021") + theme(axis.text.x = element_text(angle = 20, size = 8))
  
```

## Intake type `Owner Surrender` ##


```{r Canine Owner Surrender}
OC_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Canine") %>% 
  filter(`Intake Type` == "Owner Surrender") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

OC_bar <- OC_bar[1:10,];

ggplot(OC_bar, aes(x=Breed, y=BreedCount)) + geom_bar(stat = "identity", fill= "dark blue", color= "black", width = .60)+ labs(title = "Top 10 Canine Breeds With Intake Type as Owner Surrender", caption =  "From Years 2013 - 2021") + theme(axis.text.x = element_text(angle = 20,size = 8))
```

```{r Feline Owner Surrender}
OF_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Feline") %>% 
  filter(`Intake Type` == "Owner Surrender") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

OF_bar <- OF_bar[1:10,];

ggplot(OF_bar, aes(x=Breed, y=BreedCount)) + geom_bar(stat="identity",fill ="pink",  color= "black", width = .60) + labs(title = "Top 10 Feline Breeds With Intake Type as Owner Surrender", caption = "From 2013 - 2021") + theme(axis.text.x = element_text(angle = 20, size= 8))
```

## Intake Type `Public Assist` ##


```{r Canine Public Assist}
PC_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Canine") %>% 
  filter(`Intake Type` == "Public Assist") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

PC_bar <- PC_bar[1:10,];

ggplot(PC_bar, aes(x= Breed, y=BreedCount))+ geom_bar(stat = "identity", fill = "dark blue", color= "black", width = .60) +labs(title = "Top 10 Canine Breeds With Intake Reason as Public Assist", caption = "From Years 2013 - 2021") + theme(axis.text.x = element_text(angle = 20,size = 8))
```

```{r Feline Public Assist}
PF_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Feline") %>% 
  filter(`Intake Type` == "Public Assist") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount))

PF_bar <- PF_bar[1:10,]

ggplot(PF_bar, aes(x=Breed, y=BreedCount)) + geom_bar(stat = "identity", color = "black", fill = "pink", width = .60) + labs(title = "Top 10 Feline Breeds With Intake Type as Public Assist", caption = "From Years 2013 - 2021") + theme(axis.text.x = element_text(angle = 20,size = 8))
```


## Intake Type `Abandoned` ##


```{r Canine Abandoned}
AC_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Canine") %>% 
  filter(`Intake Type` == "Abandoned") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

AC_bar <- AC_bar[1:10,];

ggplot(AC_bar, aes(x= Breed ,y= BreedCount)) + geom_bar(stat = "identity", fill= "dark blue", color = "black", width = .30) + labs( title = "Top 10 Canine Breeds With Intake Type as Abandoned", caption = "From Years 2013 - 2021") + theme(axis.text.x=element_text(angle = 20, size = 8))
```


```{r Feline Abandoned}
AF_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Feline") %>% 
  filter(`Intake Type` == "Abandoned") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

AF_bar <- AF_bar[1:8,]

ggplot(AF_bar, aes(x=Breed, y=BreedCount)) + geom_bar(stat = "identity", fill="pink", color = "black", width = .60) +labs(title = "Top 8 Feline Breeds With Intake Type as Abandoned", caption = "From Years 2013 - 2021")+ annotate("text",x=6,y=40, label = "*Only 8 populated Feline Breeds Were Brought in Abandoned*",color = "black",angle = -20) + theme(axis.text.x = element_text(angle = 20,size = 8))
```


## Intake Type `Euthanasia Request` ##


```{r Canine Euthanasia Request}
EC_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Canine") %>% 
  filter(`Intake Type` == "Euthanasia Request") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

EC_bar <- EC_bar[1:10,];
  
  ggplot(EC_bar, aes(x=Breed,y= BreedCount)) + geom_bar(stat = "identity", fill= "dark blue", color = "black", width = .60) + labs(title = "Top 10 Canine Breeds With Intake as Euthanasia Request", caption = "From Years 2013-2021") + theme(axis.text.x = element_text(angle = 20, size = 8))
```


```{r Feline Euthanasia Request}
EF_bar <- a_df %>% 
  group_by(Breed) %>% 
  filter(Species == "Feline") %>% 
  filter(`Intake Type` ==  "Euthanasia Request") %>% 
  summarise(BreedCount = n()) %>% 
  arrange(desc(BreedCount));

EF_bar <- EF_bar[1:7,];

ggplot(EF_bar, aes(x=Breed , y= BreedCount)) + geom_bar(stat = "identity", fill="pink", color="black", width = .60) + labs(title = "Top 7 Feline Breeds With Intake Type as Euthanasia Request", caption = "From Years 2013-2021") + theme(axis.text.x = element_text(angle = 20,size = 8)) + annotate("text",x =5,y=30, label = "*Only 7 Feline Breeds Populated For Euthanasia Request*", angle = -10, size = 4)
```





